Towards a Neural Co-Processor Which Restores Movement After Stroke: Modeling a Proof-of-Concept

### 2021-11-09 paper/article.tex
Paper progress

\begin{abstract}
\textit{Objective} Brain co-processors\cite{rao.coproc} are devices which use artificial
intelligence (AI) for closed-loop neurostimulation, to shape neural activity and to bridge
injured neural circuits for targeted repair and rehabilitation. The co-processor framework
offers a flexible approach to learning closed-loop stimulation policies that optimize for
(a) specific regimes of neural activity, or (b) external task performance.
For example, it may seek to learn to stimulate the motor cortex of a stroke patient,
conditioning the stimulation on upstream visual information, aiding the patient's
attempt to grasp an object. Through the use of artificial neural
networks (ANNs) and deep learning, the co-processor co-adapts with the
neural circuit, allowing it to seek optimal stimulation policies, and adapt them
as the neural circuit changes. The results presented here demonstrate a
neural co-processor for the first time, through the use of a simulation. We
explore some of the core algorithms that may allow co-processors to successfully
learn and to adapt to non-stationarity in both the brain and sensors.
\textit{Approach} We provide the first proof-of-concept of a neural co-processor that
leverages deep learning, through the use of a simulated neural circuit
That circuit performs reach-to-grasp task, based on visual input, and is designed to
closely resemble a similar circuit in a primate brain \cite{michaels.mrnn}. We simulate
a variety of lesions by altering the model, and then demonstrate a co-processor's ability
to restore lost function through ``stimulation'' of that model. We further test the
ability of a co-processor to adapt its stimulation as the simulated brain undergoes changes.
\textit{Main results} Our simulated co-processor successfully co-adapts with the neural
circuit to accomplish the external reaching task. The co-processor framework
demonstrated here adapts to a variety of lesion types, and to ongoing changes in the
simulated brain.
\textit{Significance} The proof-of-concept here outlines a co-processor model, as well
as our approach to training it, leading to insights on how such a model may be
developed for \textit{in vivo} use. We believe this co-processor design will allow for
learning complex stimulation policies that help restore function to a stroke victim.
\end{abstract}

@@

\subsection{Simulation Overview}
We demonstrate our approach here using a simulated grasping circuit.
A detailed simulation such as this allows us to explore some of the critical details in
training a co-processor of this design. By first doing such exploration in
simulation, we are able to rapidly and cheaply iterate on our design, prior to
any \textit{in vivo} experiments. In order for the simulation to be
admissible, we need to ensure it has some properties that allow it to strongly
indicate if our design is improving in a direction that will later allow for
real deployments. Otherwise, our design may be adapting to the pecularities of
the simulation, without becoming more useful for the real world.

An example of such an approach is Bernal et al. \cite{bernal.sim}. In this work, the
authors use a simulated spiking neural network to train a stimulation agent, much
like the work we present here. Their stimulation agent sought to restore the network's
control of a simulated arm, to reach a target. As they point out, we have a limited
ability to probe a neural circuit \textit{in vivo} in order to perform learning. As
a result, we first need to design our approach through the use of an admissible
simulation. The authors in this case simulated lesions by effectively removing parts
of their simulated network, or by cutting connections between parts of the network.
Likewise here, we use an established design for an artificial neural circuit which performs
a grasping task, whose architecture and training methods were designed specifically
to result in naturalistic dynamics.

The simulated circuit, from Michaels et al. \cite{michaels.mrnn}, was trained
to resemble the grasping circuits of monkey subjects engaged in a delayed reach-to-grasp task.
Its design draws on a body of literature focused on architectures and training methods for RNNs which
seek to create artificial neural circuits that have activation dynamics similar to natural circuits,
including delayed grasping tasks \cite{susillo.mrnn}. The Michaels circuit consists of a
``modular'' vanilla RNN (mRNN), and a linear readout layer. The inputs are visual features
intended to capture the view the monkeys had during the task, i.e. 3D renderings of the same
objects which the monkeys grasped. The outputs are muscle length velocities for the should,
arm, and hand of the monkey during the trial. The natural velocities were captured with a
motion capturing glove, and the artificial neural network was trained to recapitulate those
grasping motions. Data and trained models from this work were supplied to us by the lead
author Jonathan Michaels, for the purpose of our present simulation. See Fig. \ref{fig:michaels}.

@@

The design of the circuit is intended to resemble a vision-to-grasp pipeline, essentially
representing the visual processing needed to reach the hand to the appropriate position for
the grasp, and to form the hand properly for grasping the particular shape of object. The
emergent dynamics of the artificial network's ``modules'', once trained, correspond roughly
to the AIP, F5, and M1 portions of the monkey subjects' brains. `Correspond' here, notably,
refers to the fact that the activity of the module receiving the visual inputs resembles
the AIP activity of the monkey subject from the same trials. Likewise, the second and
third modules resemble the natural activity from the monkey's F5 and M1 regions, respectively.
The emergent network dynamics show a number of other correspondences to the monkeys' natural
brain activity as well, as detailed in that paper.

@@

\subsubsection{Simulated lesions cause real world failure modes}
Simulating a brain lesion in terms of this artificial network results in error
modes that resemble a natural lesion of certain regions of a primate brain. For example, if we
alter the network by zeroing the outputs of some of the first (input, or AIP) module's neurons,
the reaching motion generally succeeds, but the finger muscle velocities show a high degree
of error - effectively implying that the subject can somewhat reach to grasp, but cannot form a
grasp appropriate for the object. That suggests that losing a portion of the cortical
machinery needed to translate or forward-propagate object shape information
to movement-related cortex can result in a reduced ability to form the hand properly for
the grasp, though positioning of the hand may still roughly succeed. Muscle spasticity of
the hand is also a common symptom of certain strokes in primates, and indeed many who
suffer from it are still able to position their hand, even while being unable to form
it properly for a grasp \cite{khanna.openloop}. In that sense, the error mode of this
simulated lesion closely resembles natural stroke symptoms. Conversely, if we simulate a
``stroke'' of a portion of the output module of the Michaels mRNN, roughly
corresponding to M1, we see a more wholesale loss of movement, affecting even the
ability to reach for the grasp. Finally, if we ``disconnect'' communications between the
F5 and M1 modules, we see a failure similar to an AIP lesion: movement is generally
achieved, but we see a disproportionate impact on hand pose.
See Fig. \ref{fig:lesion} for examples.

\begin{figure}
	\centering
	\begin{subfigure}[c]{0.62\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_trajs.pdf}
	    \caption{Example muscle trajectories during a single trial, for one hand and shoulder muscle}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.32\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_hand.pdf}
	    \caption{Proportion of loss attributable to hand muscles}
	\end{subfigure}
	\hfill
	\caption{Lesion designs which prevent forward propagation of object shape information
	         differentially impact hand pose. Loss here is an L2 distance
		 measured relative to the circuit's trajectories prior to the lesion.}
	\label{fig:lesion}
\end{figure}

@@

In our experiments below, we demonstrate our co-processor design on three types of
simulated lesions:
\begin{itemize}
	\item \textbf{Connection-based:} we prevent the forward and backward propagation of
	      information between the ``F5'' and ``M1'' modules, effectively representing a
	      severing of the connections between the two.
	\item \textbf{Output-based (AIP):} we force the output of some proportion of ``AIP''
	      neurons to zero, effectively removing them from the network. This results in a
	      loss of object shape information and, as we will see below, causes the co-processor
	      to be unable to differentiate objects.
	\item \textbf{Output-based (M1):} we force the output of some proportion of ``M1''
	      neurons to zero, causing a more holisitic loss of movement.
\end{itemize}

\subsubsection{Simulated network exhibits long running dynamics}
Just as the co-processor must adapt to the lesion, it must also adapt to the network
dynamics.

\begin{figure}
	\centering
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_single.pdf}
	    \caption{}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_dist.pdf}
	    \caption{}
	\end{subfigure}
	\hfill
	\caption{Instantaneous perturbations of a random sample (n=10) of M1 neurons at time t=200 results in long-running effects on output.}
	\label{fig:dynamics}
\end{figure}

\subsubsection{Observation model}
** Observation model
*** Spatio-temporal smoothing

@@

\subsubsection{Stimulation model}
** Stimulation design

@@

\subsection{Training the Model}
Alternate training EN and CPN.
\begin{itemize}
	\item Train an EN
		\begin{itemize}
			\item Training the EN until predictive power reaches a threshold, based on current task performance
			\item Training regime based on most recent CPN, similar CPNS (noise added to params), and white noise stimulation
			\item Single batch of data from applying current CPN set and white noise to the brain, EN fit to ``offline'' it over maining epochs
			\item Predictive power alone is not sufficient for use in training the CPN: we need to form the batch as above, otherwise training is unstable.
		\end{itemize}
	\item Train the CPN
		\begin{itemize}
			\item Backprop through the EN, effectively using the EN's prediction as ground truth
		\end{itemize}
\end{itemize}

\section{Results}

* Task performance improves drastically
* Fine tuning takes a long time
* Object classes separate
* Can co-adapt with the brain, as it changes

* Split by lesion design

* White noise and noised params
** Learning instability
** Predictive power isn't enough
** Illustrate this

* Training efficiency analysis

\section{Discussion}
* Training efficiency
* Spectrum from simple low dimensional stimulation vectors today to higher dimensional future
* Toward in vivo application

\section{Conclusion}
asdf

>>>>

\begin{abstract}
\textit{Objective} Brain co-processors\cite{rao.coproc} are devices which use artificial
intelligence (AI) for closed-loop neurostimulation, to shape neural activity and to bridge
injured neural circuits for targeted repair and rehabilitation. The co-processor framework
offers a flexible approach to learning closed-loop stimulation policies that optimize for
(a) specific regimes of neural activity, or (b) external task performance.
For example, it may seek to learn to stimulate the motor cortex of a stroke patient,
conditioning the stimulation on upstream visual information, aiding the patient's
attempt to grasp an object. Through the use of artificial neural
networks (ANNs) and deep learning, the co-processor co-adapts with the
neural circuit, allowing it to seek optimal stimulation policies, and adapt them
as the neural circuit changes. The results presented here demonstrate a
neural co-processor for the first time, through the use of a simulation. We
explore some of the core algorithms that may allow co-processors to successfully
learn and to adapt to non-stationarity in both the brain and sensors.
\textit{Approach} We provide the first proof-of-concept of a neural co-processor that
leverages deep learning, through the use of a simulated neural circuit.
That circuit performs reach-to-grasp task, based on visual input, and is designed to
closely resemble a similar circuit in a primate brain \cite{michaels.mrnn}. We simulate
a variety of lesions by altering the model, and then demonstrate our co-processor's ability
to restore lost function through ``stimulation'' of that model. We further test the
ability of our co-processor to adapt its stimulation as the simulated brain undergoes changes.
\textit{Main results} Our simulated co-processor successfully co-adapts with the neural
circuit to accomplish the external reaching task. The co-processor framework
demonstrated here adapts to a variety of lesion types, and to ongoing changes in the
simulated brain.
\textit{Significance} The proof-of-concept here outlines a co-processor model, as well
as our approach to training it, leading to insights on how such a model may be
developed for \textit{in vivo} use. We believe this co-processor design will allow for
learning complex stimulation policies that help restore function to a stroke victim.
\end{abstract}

@@

\subsection{Simulation Overview}
We demonstrate our approach here using a simulated grasping circuit.
A detailed simulation such as this allows us to explore some of the critical
architectural details and training algorithms for a co-processor of this type.
By first doing such exploration in simulation, we are able to rapidly and
cheaply iterate on our design, prior to any \textit{in vivo} experiments.
In order for the simulation to be admissible, we need to ensure it has some
properties that allow it to strongly indicate if our design is improving in
a direction that will later allow for real deployments. Otherwise, our
design may be adapting to the pecularities of the simulation, without
becoming more useful for the real world.

An example of such a simulation approach is Bernal et al. \cite{bernal.sim}. In this work, the
authors use a simulated spiking neural network to train a stimulation agent, much
like the work we present here. Their stimulation agent sought to restore the network's
control of a simulated arm, to reach a target, after a simulated lesion was applied.
As they point out, we have a limited ability to probe a neural circuit \textit{in vivo}
in order to perform learning. As a result, we first need to design our approach through
the use of an admissible simulation. The authors in this case simulated lesions by
effectively removing parts of their simulated network, or by cutting connections between
parts of the network. Likewise here, we use an established design for an artificial
neural circuit which performs a grasping task, whose architecture and training methods
were designed specifically to result in naturalistic dynamics.

The simulated circuit, from Michaels et al. \cite{michaels.mrnn}, was trained
to resemble the grasping circuits of monkey subjects engaged in a delayed reach-to-grasp task.
Its design draws on a body of literature focused on architectures and training methods for RNNs which
seek to create artificial neural circuits that have activation dynamics similar to natural circuits,
including delayed grasping tasks \cite{susillo.mrnn}. The Michaels circuit consists of a
``modular'' vanilla RNN (mRNN), and a linear readout layer. Each ``module'' consists of 100 vanilla
RNN neurons, with a non-linearity applied on the outputs. The modules are internally fully connected,
and are connected to each other sparsely (10\% connectivity). The inputs are visual features
intended to capture the view the monkeys had during the task, specifically VGGNet features from 3D
renderings of the same objects which the monkeys grasped. The outputs are muscle length
velocities for the should, arm, and hand of the monkey during the trial. The natural
velocities were captured with a motion capturing glove, and the artificial neural network
was trained to recapitulate those grasping motions. Data and trained models from this work
were supplied to us by the lead author Jonathan Michaels, for the purpose of our present
simulation. See Fig. \ref{fig:michaels}.

@@

The design of the circuit is intended to resemble a vision-to-grasp pipeline, essentially
representing the visual processing needed to reach the hand to the appropriate position for
the grasp, and to form the hand properly for grasping the particular shape of object. The
emergent dynamics of the artificial network's ``modules'', once trained, correspond roughly
to the AIP, F5, and M1 portions of the monkey subjects' brains. `Correspond' here, notably,
refers to the fact that the activity of the module receiving the visual inputs resembles
the AIP activity of the monkey subject from the same trials. Likewise, the second and
third modules resemble the natural activity from the monkey's F5 and M1 regions, respectively.
The emergent network dynamics show a number of other correspondences to the monkeys' natural
brain activity as well, as detailed in the paper.

@@

\subsubsection{Simulated lesions cause real world failure modes}
Simulating a brain lesion in terms of this artificial network results in error
modes that resemble a natural lesion of certain regions of a primate brain. For example, if we
alter the network by zeroing the outputs of some of the first (input, or AIP) module's neurons,
the reaching motion generally succeeds, but the finger muscle velocities show a high degree
of error - effectively implying that the subject can somewhat reach to grasp, but cannot form a
grasp appropriate for the object. That suggests that losing a portion of the cortical
machinery needed to translate or forward-propagate object shape information
to movement-related cortex can result in a reduced ability to form the hand properly for
the grasp, though positioning of the hand may still roughly succeed. Muscle spasticity of
the hand is also a common symptom of certain strokes in primates, and indeed many who
suffer from it are still able to position their hand, even while being unable to form
it properly for a grasp \cite{khanna.openloop}. In that sense, the error mode of this
simulated lesion closely resembles natural stroke symptoms, though the simulation is not
intended to constitute a physical model of a lesion, i.e. to directly explain the
connection between hand spasticity and the lesion. Conversely, if we
lesion of a portion of the output module of the Michaels mRNN, roughly
corresponding to M1, we see a more wholesale loss of movement, affecting even the
ability to reach for the grasp. Finally, if we ``disconnect'' communications between the
F5 and M1 modules, we see a failure similar to an AIP lesion: movement is generally
achieved, but we see a disproportionate impact on hand pose.
See Fig. \ref{fig:lesion} for examples.

\begin{figure}[h]
	\centering
	\begin{subfigure}[c]{0.62\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_trajs.pdf}
	    \caption{Example muscle trajectories during a single trial, for one hand and shoulder muscle}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.32\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_hand.pdf}
	    \caption{Proportion of loss attributable to hand muscles}
	\end{subfigure}
	\hfill
	\caption{Lesion designs which prevent forward propagation of object shape information
	         differentially impact hand pose. Loss here is an L2 distance
		 measured relative to the circuit's trajectories prior to the lesion.}
	\label{fig:lesion}
\end{figure}

@@

In our experiments below, we demonstrate our co-processor design on three types of
simulated lesions:
\begin{itemize}
	\item \textbf{Output-based (AIP):} we force the output of some proportion of ``AIP''
	      neurons to zero, effectively removing them from the network. This results in a
	      loss of object shape information and, as we will see below, causes the co-processor
	      to be unable to differentiate objects.
	\item \textbf{Connection-based:} we prevent the forward and backward propagation of
	      information between the ``F5'' and ``M1'' modules, effectively representing a
	      severing of the connections between the two.
	\item \textbf{Output-based (M1):} we force the output of some proportion of ``M1''
	      neurons to zero, causing a more holisitic loss of movement.
\end{itemize}

\subsubsection{Simulated network exhibits long running dynamics}
Just as the co-processor must adapt to the lesion, it must also adapt to the dynamics of
the network it is stimulating. Natural neural networks as well as our stimulated network
exhibit long running dynamics, which our CPN and EN must account for in their learning.
A perturbation of a network (i.e. due to stimulation) will cause changes in neuron
activations long after the stimulation is applied, sometimes far from the site of
stimulation. Our simulated grasping circuit exhibits the same behavior.

To illustrate this, suppose we applied a small, one-time, instantaneous perturbation
of the hidden state of 10 randomly chosen neurons in the output module at some point
in time during a trial. If we repeat that experiment many times, we can see what the
distribution of long- running effects tends to look like.

In Fig. \ref{fig:dynamics} we can see that even a single, one-time perturbation in
the network has effects dozens of time steps later. Our co-processor will need to
account for these, since stimulation is intended to cause perturbations in a network.
To successfully aid in grasping, the co-processor will need to account for these
dynamics. As we will show in the next subsection, the problem the co-processor faces
is in-fact even harder than this, due to a stimulation model that includes both
spatial and temporal smoothing.

\begin{figure}[h]
	\centering
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_single.pdf}
	    \caption{Example trial}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.48\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{perturbe_dist.pdf}
	    \caption{Distribution of effects across n=1000 samples}
	\end{subfigure}
	\hfill
	\caption{Instantaneous perturbations of a random sample (n=10) of M1
	         neurons at time t=200 results in long-running effects on output.}
	\label{fig:dynamics}
\end{figure}

\subsubsection{Stimulation model}
TODO: overview of spatio-temporal smoothing, connect to figure.

\subsubsection{Observation model}
TODO: overview of spatial / gaussian convolution concept, connect to $fig:stim_and_obs$.

@@

@@

\subsection{Training the Co-Processor}
Alternate training EN and CPN.
\begin{itemize}
	\item Train an EN
		\begin{itemize}
			\item Training the EN until predictive power reaches a threshold, based on current task performance
			\item Training regime based on most recent CPN, similar CPNS (noise added to params), and white noise stimulation
			\item Single batch of data from applying current CPN set and white noise to the brain, EN fit to ``offline'' it over maining epochs
			\item Predictive power alone is not sufficient for use in training the CPN: we need to form the batch as above, otherwise training is unstable.
			\item White noise and noised params
			\item ** Learning instability
			\item ** Predictive power isn't enough
			\item ** Illustrate this
		\end{itemize}
	\item Train the CPN
		\begin{itemize}
			\item Backprop through the EN, effectively using the EN's prediction as ground truth
		\end{itemize}

	\item Data recycling/reuse
\end{itemize}

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{pseudocode.png}
\caption{Pseudocode of the Training Method}
\centering
\label{fig:pseudocode}
\end{figure}

\section{Results}
\begin{itemize}
	\item Task performance improves drastically.
		\begin{itemize}
			\item Fig: normalized task loss vs CPN training epoch
			\item Fig: normalized task loss vs training epoch
			\item Fig: normalized task loss vs training epoch
			\item Split by experiment
		\end{itemize}
	\item Fine tuning takes a long time
	\item Object classes separate. Fig: sep vs experiment vs epoch
	\item Can co-adapt with the brain, as it changes
	\item Training efficiency analysis. Show num epochs with and without recycling.
\end{itemize}

\section{Discussion and Conclusion}
\begin{itemize}
	\item Training efficiency
	\item Spectrum from simple low dimensional stimulation vectors today to higher dimensional future
	\item Toward in vivo application
\end{itemize}
